---
title: |
  A new simplified manual tour, with examples in mathematica
type: Short Technical Note
author:
  - name: Alex Aumann
    affil: a
    email: aaum0002@student.monash.edu
  - name: German Valencia
    affil: a
    email: german.valencia@monash.edu
  - name: Ursula Laa
    affil: b
    email: ursula.laa@boku.ac.at
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Something here
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE,
  warning = FALSE)
```

```{r libraries}
# Load libraries
library(tourr)
library(plotrix)
```

%\doublespacing
# Introduction

<!-- Background -->

# Manual tour {#sec:method}

```{r manualsequence, outwidth="80%", width=12, height=8, fig.cap="Sequence of projections where V3 contribution is changed."}
source("../code/manual.R")

# Base plot set up
plot.new()
par(pty="s", xaxt="n", yaxt="n", bty="n",
      omi=c(0,0,0,0), mar=c(0,0,0,0),
    mfrow=c(3,4))

# Create manual tour example
p <- 4
d <- 2
set.seed(24)
A <- matrix(runif(p*d, min=-1), ncol=d, byrow=TRUE)
colnames(A) <- c("P1", "P2")
rownames(A) <- paste0("V", 1:p)
A <- tourr::orthonormalise(A)
plot_basis(A)
  
# Now set a new position
vchange <- 3
eps <- 0.02

# Iterate and save images
for (i in 1:11) {
  flname <- paste0("mtour_plots/mtour", i, ".png")
  A[vchange,] <- A[vchange,]*(1-i*eps)
  A <- tourr::orthonormalise(A)
  #png(flname, width=250, height=250)
  plot_basis(A)
  #dev.off()
}
```

## Background

## New definition





# Implementation {#sec:implementation}



# Applications {#sec:examples}


# Discussion {#sec:discussion}



# Acknowledgements {-}

The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr]. 

# Supplementary material {-}

The source material and animated gifs for this paper are available at 
